---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# Disclaimer

This software is part of the DATA501 2024 course taught a Victoria University in Wellington, and is intended for educational purposes only. 

# OutlierDetection

<!-- badges: start -->
<!-- badges: end -->

The goal of `OutlierDetection` is to calculate measure of influence which helps to user to detect potential outliers.




## Installation 

You can install the development version of `OutlierDetection` like so:

``` r
devtools::install_github("DavidHuijser/DATA501OutlierDetection",build_vignettes=TRUE )
```

Once you successfully installed `OutlierDetection` you can add it to your library using the following command: 

``` r
library(OutlierDetection)  
```

## Manual and Vignette 

This allows you to view the general vignette using command:

```r
browseVignettes("OutlierDetection")
```

The result of the might be dependent on the specific setup of your computer and `R` environment. 

The function available to the user is `InfluenceMeasure`. In the next section you will see an example, however if you prefer you can find the manual using either 

```r
?InfluenceMeasure
```
or 

```r
help(InfluenceMeasure)
```

in the `R` console. 


## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(OutlierDetection)
slope <-  4
offset <- -3
sds <- 7
x <- seq(from = 1, to = 10, length.out= 100)

y <- slope*x + offset

# Add regular noise
noise <- rnorm(length(x),sd=2)
y <- y + noise

# generate number of outliers
num <- 3
s <- sample(length(x), num)
y[s] <- y[s] +  rnorm(num, sd=sds) + runif(num, -sds,sds)

# Create plot
df <- data.frame(x,y)
lin_model <- lm(y~x)


InfluenceMeasure(lin_model, measure="Cooks", output = "values")
InfluenceMeasure(lin_model, measure="Cooks", output = "plot")


```



